---
title: "UFO code"
author: "Shealagh Brown & Sam Zimpfer"
date: "2025-03-20"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(gapminder, socviz, tidyverse, grid, ggthemes,
               usmap, maps, statebins, viridis, leaflet, lubridate, broom, GGally, rpart, rpart.plot, FNN, class, caret) 
theme_bw()
```

---


## Cleaning the data

The variables used are:
**datetime**: the date and time of sighting in year, month, day, hours, minutes, seconds format. \
**city**: city where sighitng occurred \
**state**: state where sighting occurred \
**country**: country where sighting occurred \
**seconds**: duration of sightings in seconds \
**latitude**: latitude of sighting \
**longitude**: longitude of sighing \
**year**: year when sighing occurred \
**month**: month when sighting occurred \

```{r clean data}
ufo_raw <- read.csv("ufo_sightings_scrubbed.csv")

#convert date time to ymd_hms format
ufo_raw$datetime <- ymd_hms(ufo_raw$datetime)

#cleaning data
ufo_raw|>
  # keep duration in seconds form only
   mutate(seconds = duration..seconds.,
          # split datetime data into a year and a month column
          year = year(datetime),
          month = month.name[month(datetime)])|>
  # select relevant columns
  select(datetime, city, state, country, seconds,latitude,longitude, year, month)|>
    filter(seconds <= 40000) |> 
    # filter out badly formatted entries that could cause NA's during the following conversion
    filter(grepl("^[-]?[0-9.]+$", latitude),
           grepl("^[-]?[0-9.]+$", seconds)) -> ufo

  
ufo$latitude <- as.numeric(ufo$latitude) #changing lat to numeric
ufo$seconds <- as.numeric(ufo$seconds) #changing seconds to numeric

# drop any remaining NA entries
ufo <- ufo |> drop_na(longitude, latitude, seconds)

```

---


# Data Summary

The first question we wanted to answer is where do most UFO sightings occur. 

We created histograms of sightings based on longitude and latitude.
```{r pressure, longitude & latitude }
# histogram for longitude
long_hist <- ggplot(
  data = ufo,
  mapping = aes( x= longitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black", 
                 bins = 40)+
  labs(
    title = "Number of sightings per longitude",
    x = "Longitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-200, 200, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))


# histogram for latitude
lat_hist <- ggplot(
  data = ufo,
  mapping = aes( x= latitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black",
                 bins = 40)+
  labs(
    title = "Number of sightings per lattitude",
    x = "Latitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-100, 100, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))

#display both graphs
long_hist
lat_hist
```

---


Next, we overlayed the coordinates of the sightings on a map of the world to better visualize.
```{r , map }
#createing scatter plot ontop of world map
map <- map_data("world")

map_plot <- ggplot()+
  geom_polygon(data = map,
               mapping = aes(x= long,
                             y = lat,
                             group = group),
               fill = "darkgreen")+
  geom_point(data = filter(ufo, seconds < 40000),
             mapping = aes(x = longitude,
                           y = latitude,
                           color = seconds),
             size = 0.3
           )+
  labs(
    x = "Longitude",
    y = "Latitude"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
  ylim(-60,90)

#displaying map plot 
map_plot
```


---


## Bar graph of sightings per year
```{r, bar graph of sightings per year}

#creating plot of sightings per year
sightings_year<- ggplot( data = ufo,
        mapping = aes(x = year))+
  geom_histogram(fill = "hotpink",
                 color = "black",
                 bins = 20)+
  labs(
    x = "Year",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))

sightings_year
```

---


## Small Multiples Graph of Number of Sightings Per Month Since Year 2000
```{r, small multiples}
# create table of year, month, number of sightings
summary <- ufo |> filter(year >= 2000) |> group_by(year, month) |> summarise(count = n())

# order data by natural month ordering
summary$month <- factor(summary$month, levels = month.name)
summary <- summary |> arrange(year, month)

# create small multiples graph
months_by_year <- ggplot(data = summary,
       mapping = aes(x = month,
                     y = count,
                     group = 1)) +
  geom_line() +
  labs(x = "Month",
       y = "Number of Sightings") +
  scale_y_continuous(expand = c(0, 0, 0.05, 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ year)

#display graph
months_by_year
```



#Machine Learning

## Setting up linear regression models
```{r,  models}
# create several models to explain seconds in terms of latitude and/or longitude
lat_lm<- lm(
  formula = seconds ~ latitude,
  data = ufo 
)
  
long_lm<- lm(
  formula = seconds ~ longitude,
  data = ufo 
)
  
lat_long_lm<- lm(
  formula = seconds ~ latitude + longitude,
  data = ufo 
)

# create table to compare performance of the different models
bind_rows(
  glance(lat_lm),
  glance(long_lm),
  glance(lat_long_lm)
  ) |> 

  mutate(
    explanatories = c(as.character(formula(lat_lm))[3],
                      as.character(formula(long_lm))[3],
                      as.character(formula(lat_long_lm))[3])
  ) |> 
select(explanatories, r.squared, adj.r.squared, sigma)  


```

---


## Residual Plots for Duration of Sightings vs Longitude and Lattitude
```{r, residual plots}
# define function to show a residual plot of a specified regression model
resid_plot_print <- function(model, title) {
  augment_columns(
  x = model, 
  data = ufo
) |> 
  
  ggplot(
    mapping = aes(
      x = .fitted,
      y = .resid
    )
  ) + 
  
  geom_point(alpha = 0.5) + 
  
  geom_hline(
    yintercept = 0,
    color = "red",
    linewidth = 1
  ) + 
  
  labs(
    x = "Predicted Values",
    y = "Residuals",
    title = paste("Residual Plot for" , title , "Model"
    )) }

#display residual plots
resid_plot_print(lat_lm, "Latitude")
resid_plot_print(long_lm, "Longitude" )
resid_plot_print(lat_long_lm, "Longitude and Latitude" )

```






## kNN regression

``` {r ,  kNNreg }
#creating normalizing function
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
#creating standardizing function
standardize <- function(x) {
  return((x - mean(x)) / sd(x))
}
#normalizing explanatory variables
ufo_norm <- 
  ufo |> 
  select(longitude, latitude, seconds) |> 
  mutate(
    across(
      .cols = latitude:longitude,
      .fns = normalize
    )
  )
#standardizing explanatory variables 
ufo_stan <- 
  ufo |> 
  select(longitude, latitude, seconds) |> 
  mutate(
    across(
      .cols = latitude:longitude,
      .fns = standardize
    )
  )


k <- 1:100


# kNN with normalizing data

#making data frame for normalized stats
fit_stats_norm <- 
  data.frame(k = k,
         R2 = rep(-1, length(k)),
         MAE = rep(-1, length(k)))
#creating kNN loop
for (i in 1:length(k)) {
  norm_knn <- 
    knn.reg(
      train = ufo_norm,
      y = ufo$seconds,
      k = k[i]
    )
  
  fit_stats_norm[i, "R2"] <- norm_knn$R2Pred
  fit_stats_norm[i, "MAE"] <- (ufo$seconds - norm_knn$pred) |> abs() |> mean()
}

#plotting normalized kNN fit stats
fit_stats_norm |> 
  pivot_longer(
    cols = R2:MAE,
    names_to = "fit_stat",
    values_to = "fit"
  ) |> 
  
  ggplot(mapping = aes(x = k,
                       y = fit,
                       color = fit_stat)) + 
  geom_line(show.legend = F) + 
  facet_wrap(facets = ~ fit_stat,
             scales = "free_y",
             ncol = 1)+
  labs(title = "Fit Statistics for Normalized data")


# kNN standardizing

#creating data frame for standardized stats
fit_stats_stan <- 
  data.frame(k = k,
         R2 = rep(-1, length(k)),
         MAE = rep(-1, length(k)))
#making kNN loop
for (i in 1:length(k)) {
  stan_knn <- 
    knn.reg(
      train = ufo_stan,
      y = ufo$seconds,
      k = k[i]
    )
  
  fit_stats_stan[i, "R2"] <- stan_knn$R2Pred
  fit_stats_stan[i, "MAE"] <- (ufo$seconds - stan_knn$pred) |> abs() |> mean()
}


#plotting standarsized kNN fit stats
fit_stats_stan |> 
  pivot_longer(
    cols = R2:MAE,
    names_to = "fit_stat",
    values_to = "fit"
  ) |> 
  
  ggplot(mapping = aes(x = k,
                       y = fit,
                       color = fit_stat)) + 
  geom_line(show.legend = F) + 
  facet_wrap(facets = ~ fit_stat,
             scales = "free_y",
             ncol = 1)+
  labs(title = "Fit Statistics for Standardized data")

#combining for comparison
fit_stats_combined <- 
  bind_rows("stan" = fit_stats_stan,
            "norm" = fit_stats_norm,
            .id = "rescale")

#finding best k value
fit_stats_combined |>
  filter(
    R2 == max(R2) | MAE == min(MAE)
    )


#the running kNN with best k value and rescale method
norm_knn <-
  knn.reg(
    train = ufo_norm,
    y = ufo$seconds,
    k = 1
  )

```


## Making the kNN classification
``` {r ,  kNNclass }

#creating kNN classification with normalized data
knn_norm_class <- 
  knn(
    train = ufo_norm |> select(latitude:longitude),
    test = ufo_norm |> select(latitude:longitude),
    cl = ufo$month,
    k = 1
  )
#creating kNN classification with standardized data
knn_stan_class <- 
  knn(
    train = ufo_stan |> select(latitude:longitude),
    test = ufo_stan |> select(latitude:longitude),
    cl = ufo$month,
    k = 1
  )

#combinging to get best k value
knn_acc <-
  tibble(k = 1:10,
         norm_acc = rep(-1, length(k)),
         stan_acc = rep (-1, length(k)))

#running through best k values 
for ( i in 1:nrow(knn_acc)){
  knn_norm <- 
    knn.cv(train = ufo_norm,
           cl = ufo$month,
           k = knn_acc[i , 1])
 knn_acc[i , "norm_acc"] <- mean(knn_norm == ufo$month)
 knn_stan <- 
    knn.cv(train = ufo_stan,
           cl = ufo$month,
           k = knn_acc[i , 1])
 knn_acc[i , "stan_acc"] <- mean(knn_stan == ufo$month)
}

#graphing results of loop 
knn_acc |>
  pivot_longer(
    cols = norm_acc:stan_acc,
    names_to = "rescale",
    values_to = "accuracy"
  )|>
  mutate(rescale = str_remove_all(rescale, "_acc"))|>
  ggplot(mapping = aes( x = k,
                        y = accuracy,
                        color = rescale))+
  geom_line() + 
  scale_y_continuous(labels = scales::label_percent())

#searching for best k values
knn_acc |>
  pivot_longer(cols = -k,
               names_to = "rescale_method",
               values_to = "accuracy")|>
  slice_max(accuracy)

#creating confusion matrix for the kNN classification
table(actual = ufo$month,
      predicted = knn_classification)|>
  confusionMatrix(positive = "F")

```

``` {r classification tree}
# creating full classification tree
tree_full<-
  rpart(
    formula = month ~ longitude + latitude,
    data = ufo,
    method = "class",
    parms = list(split = "information"),
    minsplit = 2,
    minbucket = 1,
    cp = -1
  )

#finding xerror cut off
tree_full$cptable |>
  data.frame()|>
  slice_min(xerror, n = 1)|> #finding row with smallest xerror
  mutate(xcutoff = xerror + xstd) |> #calculate cut off value
  pull(xcutoff) ->xcutoff
  

#use xcutoff to find corresponding cp value 
tree_full$cptable|>
  data.frame()|>
  filter(xerror < xcutoff)|> #keeping rows with x error less than x cut off
  slice(1)|>
  pull(CP) -> cp_cutoff #saving the value


# Prune the tree
prune(tree = tree_full,
      cp = cp_cutoff) -> pruned_tree

#plotting the pruned tree
rpart.plot(
  x = pruned_tree,
  type = 5,
  extra = 102
)

#displaying rules for the tree
rpart.rules(
  x=pruned_tree,
  extra = 4
)

#finding variable importance for the tree
varImp(object = pruned_tree)

#summary of the tree(not included in final paper due to length)
summary(object = pruned_tree,
        digits = 3)
```


