---
title: "UFO"
author: "Shealagh Brown & Sam Zimpfer"
date: "2025-03-20"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(gapminder, socviz, tidyverse, grid, ggthemes,
               usmap, maps, statebins, viridis, leaflet, lubridate, broom, GGally, rpart, rpart.plot, FNN) 
theme_bw()
```

## Introduction

## Cleaning the data

The code chunk below reads in our data about UFO sightings. We had to convert the datetime variable to year, month, date format. Then we created a new data frame containing only necessary variables for our purposes. Latitude had to then be convered into a numeric variable.

The variables used are:
**datetime**: the date and time of sighting in year, month, day, hours, minutes, seconds format. \
**city**: city where sighitng occurred \
**state**: state where sighting occurred \
**country**: country where sighting occurred \
**seconds**: duration of sightings in seconds \
**latitude**: latitude of sighting \
**longitude**: longitude of sighing \
**year**: year when sighing occurred \
**month**: month when sighting occured \

```{r cars}
ufo_raw <- read.csv("ufo_sightings_scrubbed.csv")

#convert date time to ymd_hms format
ufo_raw$datetime <- ymd_hms(ufo_raw$datetime)

#cleaning data
ufo_raw|>
   mutate(seconds = duration..seconds.,
          year = year(datetime), #create year column
          month = month.name[month(datetime)])|> #create month column and convert to name  
  select(datetime, city, state, country, seconds,latitude,longitude, year, month)|>
    filter(seconds <= 40000) -> ufo


  
ufo$latitude <- as.numeric(ufo$latitude) #changing lat to numeric
ufo$seconds <- as.numeric(ufo$seconds) #changing seconds to numeric


ufo <- ufo |> drop_na(longitude, latitude, seconds)

```

## Data Summary 

The first question we wanted to answer is where do most UFO sightings occur. 

We created graphs of sighingts based on longitude and latitudue 
```{r pressure, longitude & latitude }
long_hist <- ggplot(
  data = ufo,
  mapping = aes( x= longitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black", 
                 bins = 40)+
  labs(
    title = "Number of sightings per longitude",
    x = "Longitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-200, 200, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))

lat_hist <- ggplot(
  data = ufo,
  mapping = aes( x= latitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black",
                 bins = 40)+
  labs(
    title = "Number of sightings per lattitude",
    x = "Latitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-100, 100, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))

long_hist
lat_hist
```





---


## Scatter plot of sighting locations 
```{r , map }
map <- map_data("world")

map_plot <- ggplot()+
  geom_polygon(data = map,
               mapping = aes(x= long,
                             y = lat,
                             group = group),
               fill = "darkgreen")+
  geom_point(data = filter(ufo, seconds < 40000),
             mapping = aes(x = longitude,
                           y = latitude,
                           color = seconds),
             size = 0.3
           )+
  labs(
    x = "Longitude",
    y = "Latitude"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
  ylim(-60,90)

map_plot
```

### Observations
From the histograms, we noticed that most ufo sightings are concentrated around -120 through -70 longitude and 30 through 50 latitudes. When longitude and latitude are graphed against each other and over layed with a map of the word we can see that these values correspond with a high volume of sightings in the United States. This graph also indicates other potential hot spots of sightings, specifically Europe and the eastern coast of Australia. 


---


## Bar graph of sightings per country
```{r, bar graph of countries and cities}

modified_data = ufo |> filter(country != "")

ggplot(data = modified_data,
       mapping = aes(x = country))+
  geom_bar(fill = "blue4",
           color = "black",
           na.rm = T)+
scale_y_log10(expand = c(0, 0, 0.05, 0))+
  labs(title = "Total number of sightings per country",
       x = "Country",
       y = "Count (log10 scale)")+
scale_x_discrete(labels = c("Australia", "Canada", "Denmark", "Great Britian", "United States"))
  
         
```

From this graph we can determine that the United States is the country with the most sightings. There is a signficantly higher amount of sightings in the U.S. as opposed to the other countries as it is the only country with greater than 10,000 sightings. Of the five highest countries with sighitngs, Denmark has the lowest amount of sightings. They have just around 100 sighitngs reported. 

---


## Bar graph of sightings per year
```{r, bar graph of sightings per year}
ggplot( data = ufo,
        mapping = aes(x = year))+
  geom_histogram(fill = "hotpink",
                 color = "black",
                 bins = 20)+
  labs(
    x = "Year",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))
```
### Observations
UFO sightings were relatively infrequent in the early 1900's. Sightings gradually started to increase during the 1950's and then there was a sharp increase at the end of the 1990's and into the 2000's. 


```{r, bar graph of sightings per month}
by_month <- ggplot( data = ufo,
        mapping = aes(x = month))+
  geom_bar(fill = "purple",
                 color = "black")+
  labs(
    x = "Month",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
theme(
  axis.text = element_text( angle = 90
     )
)
by_month
```
### Observations
Over the years, UFO sightings have been most common during June, July, and August. The fewest occurances happend during February. 



---


## Small Multiples Graph of Number of Sightings Per Month Since Year 2000
```{r, small multiples}
# create table of year, month, number of sightings
summary <- ufo |> filter(year >= 2000) |> group_by(year, month) |> summarise(count = n())

# order data by natural month ordering
summary$month <- factor(summary$month, levels = month.name)
summary <- summary |> arrange(year, month)

ggplot(data = summary,
       mapping = aes(x = month,
                     y = count,
                     group = 1)) +
  geom_line() +
  labs(x = "Month",
       y = "Number of Sightings") +
  scale_y_continuous(expand = c(0, 0, 0.05, 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ year)
```

### Observations
When we compare how sightings per month has changed over time we see that generally the amount of sighitngs has increased. Before 2009 sightinga are generally conisistent over the year. In 2009 we can see the first real peak during the summer. This trend seems to be reflected in the following years since then. 

#Machine Learning

We want to conduct linear regression to determine the effect that longitude and latitude have on the durration of sighitngs. Before we create a model we will check for 

---


## Setting up linear regression models
```{r,  models}
lat_lm<- lm(
  formula = seconds ~ latitude,
  data = ufo 
)
  
long_lm<- lm(
  formula = seconds ~ longitude,
  data = ufo 
)
  
lat_long_lm<- lm(
  formula = seconds ~ latitude + longitude,
  data = ufo 
)

bind_rows(
  glance(lat_lm),
  glance(long_lm),
  glance(lat_long_lm)
  ) |> 

  mutate(
    explanatories = c(as.character(formula(lat_lm))[3],
                      as.character(formula(long_lm))[3],
                      as.character(formula(lat_long_lm))[3])
  ) |> 
select(explanatories, r.squared, adj.r.squared, sigma)  


```
## Residual Plots for Duration of Sightings vs Longitude and Lattitude
```{r, residual plots}
resid_plot_print <- function(model) {
  augment_columns(
  x = model, 
  data = ufo
) |> 
  
  ggplot(
    mapping = aes(
      x = .fitted,
      y = .resid
    )
  ) + 
  
  geom_point(alpha = 0.5) + 
  
  geom_hline(
    yintercept = 0,
    color = "red",
    linewidth = 1
  ) + 
  
  labs(
    x = "Predicted",
    y = "Residuals"
  )}

resid_plot_print(lat_lm)
resid_plot_print(long_lm)
resid_plot_print(lat_long_lm)

```



### Observations


---


## Setting up kNN regression- Finding the best k

``` {r finding the best k}
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

standardize <- function(x) {
  return((x - mean(x)) / sd(x))
}

ufo_norm <- 
  ufo |> 
  select(longitude, latitude, seconds) |> 
  mutate(
    across(
      .cols = latitude:longitude,
      .fns = normalize
    )
  )

ufo_stan <- 
  ufo |> 
  select(longitude, latitude, seconds) |> 
  mutate(
    across(
      .cols = latitude:longitude,
      .fns = standardize
    )
  )


k <- 1:10


# try with normalizing
fit_stats_norm <- 
  tibble(k = k,
         R2 = rep(-1, length(k)),
         MAE = rep(-1, length(k)))

for (i in 1:length(k)) {
  norm_knn <- 
    knn.reg(
      train = ufo_norm,
      y = ufo$seconds,
      k = k[i]
    )
  
  fit_stats_norm[i, "R2"] <- norm_knn$R2Pred
  fit_stats_norm[i, "MAE"] <- (ufo$seconds - norm_knn$pred) |> abs() |> mean()
}

fit_stats_norm

fit_stats_norm |> 
  pivot_longer(
    cols = R2:MAE,
    names_to = "fit_stat",
    values_to = "fit"
  ) |> 
  
  ggplot(mapping = aes(x = k,
                       y = fit,
                       color = fit_stat)) + 
  geom_line(show.legend = F) + 
  facet_wrap(facets = ~ fit_stat,
             scales = "free_y",
             ncol = 1)


# try with standardizing
fit_stats_stan <- 
  tibble(k = k,
         R2 = rep(-1, length(k)),
         MAE = rep(-1, length(k)))

for (i in 1:length(k)) {
  stan_knn <- 
    knn.reg(
      train = ufo_stan,
      y = ufo$seconds,
      k = k[i]
    )
  
  fit_stats_stan[i, "R2"] <- stan_knn$R2Pred
  fit_stats_stan[i, "MAE"] <- (ufo$seconds - stan_knn$pred) |> abs() |> mean()
}

fit_stats_stan

fit_stats_stan |> 
  pivot_longer(
    cols = R2:MAE,
    names_to = "fit_stat",
    values_to = "fit"
  ) |> 
  
  ggplot(mapping = aes(x = k,
                       y = fit,
                       color = fit_stat)) + 
  geom_line(show.legend = F) + 
  facet_wrap(facets = ~ fit_stat,
             scales = "free_y",
             ncol = 1)

fit_stats_combined <- 
  bind_rows("stan" = fit_stats_stan,
            "norm" = fit_stats_norm,
            .id = "rescale") 

fit_stats_combined |>
  filter(
    R2 == max(R2) | MAE == min(MAE)
    
  )

norm_knn <-
  knn.reg(
    train = ufo_norm,
    y = ufo$seconds,
    k = 1
  )

```

### Observations
* The best way to go about kNN regression is to use a standardized dataset with a k value of 2. This is true because this gives the highest R^2 value of 0.8.

---


## Making the kNN regression
``` {r making the regression tree}

```

``` {r regression tree cont.}

```

### Observations
* Using kNN regression proved to be 75.02% accurate in predicting the duration of sighting based on longitude and latitude.
* Based on an R^2 value of 0.7502
