---
title: "UFO"
author: "Shealagh Brown & Sam Zimpfer"
date: "2025-03-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(gapminder, socviz, tidyverse, grid, ggthemes,
               usmap, maps, statebins, viridis, leaflet, lubridate) 
theme_bw()
```

## Introduction

## Cleaning the data

The code chunk below reads in our data about UFO sightings. We had to convert the datetime variable to year, month, date format. 

```{r cars}
ufo_raw <- read.csv("ufo_sightings_scrubbed.csv")

#convert date time to ymd_hms format
ufo_raw$datetime <- ymd_hms(ufo_raw$datetime)

#cleaning data
ufo<- ufo_raw|>
   mutate(seconds = duration..seconds.,
          year = year(datetime), #create year column
          month = month.name[month(datetime)])|> #create month column and convert to name 
  select(datetime, city, state, country, seconds,latitude,longitude, year, month)
  

 

ufo$latitude <- as.numeric(ufo$latitude) #chaning lat to numeric


```

## Longitude and latitude histogram
```{r pressure, longitude & latitude }
long_hist <- ggplot(
  data = ufo,
  mapping = aes( x= longitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black", 
                 bins = 40)+
  labs(
    title = "Number of sightings per longitude",
    x = "Longitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-200, 200, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))

lat_hist <- ggplot(
  data = ufo,
  mapping = aes( x= latitude
  )
)+
  geom_histogram(fill = "olivedrab", 
                 color = "black",
                 bins = 40)+
  labs(
    title = "Number of sightings per lattitude",
    x = "Latitude",
    y = "Number of Sightings"
  )+
  scale_x_continuous(breaks = seq(-100, 100, 20))+
  scale_y_continuous(expand = c(0, 0, 0.05, 0))

long_hist
lat_hist
```

### Observations
* Longitudes of sightings appear to be concentrated around -120 through -70 (Number of sightings per longitude)
* Lattitude of sightings appear to be concentrated around 30 through 50 (Number of sightings per lattitude)

---


## Scatter plot of sighting locations 
```{r , map }
map <- map_data("world")

map_plot <- ggplot()+
  geom_polygon(data = map,
               mapping = aes(x= long,
                             y = lat,
                             group = group),
               fill = "darkgreen")+
  geom_point(data = ufo,
             mapping = aes(x = longitude,
                           y = latitude),
             size = 0.3)+
  labs(
    x = "Longitude",
    y = "Latitude"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
  ylim(-60,90)

map_plot
```

### Observations
* Most of the sightings are concentrated in North America and in Europe

---


## Bar graph of sightings per country
```{r, bar graph of countries and cities}

modified_data = ufo |> filter(country != "")

ggplot(data = modified_data,
       mapping = aes(x = country))+
  geom_bar(fill = "blue4",
           color = "black",
           na.rm = T)+
scale_y_log10(expand = c(0, 0, 0.05, 0))+
  labs(title = "Total number of sightings per country",
       x = "Country",
       y = "Count (log10 scale)")
  
         
```

### Observations
* This graph supports the fact that the US, Canada, and Great Britain contain most of the sightings

---


## Bar graph of sightings per year
```{r, bar graph of sightings per year}
ggplot( data = ufo,
        mapping = aes(x = year))+
  geom_histogram(fill = "hotpink",
                 color = "black",
                 bins = 20)+
  labs(
    x = "Year",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))

by_month <- ggplot( data = ufo,
        mapping = aes(x = month))+
  geom_bar(fill = "purple",
                 color = "black")+
  labs(
    x = "Month",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
theme(
  axis.text = element_text( angle = 90
     )
)
```

### Observations
* Sightings increased dramatically around late 1990s and early 2000s, then seemed to remain relatively constant after 2010

---


```{r, small multiples}
ufo_2000<- ufo|>
  filter(year >= 2000)

ggplot( data = ufo_2000,
        mapping = aes(x = month))+
  geom_bar(fill = "purple",
                 color = "black")+
  labs(
    x = "Month",
    y = "Number of Sightings"
  )+
scale_y_continuous(expand = c(0, 0, 0.05, 0))+
theme(
  axis.text = element_text( angle = 90
     )
)+
  facet_wrap(~ year)
  
         
```
